The Design, Semantics, and Implementation of the Synchronous Reactive Language \CEU

### 2015-10-10 paper.tex
Subfiguras (vs left/right)

Regarding the similarities, Figure~\ref{lst.abro} shows side-by-side the 
implementations in Esterel and \CEU for the following control 
specification~\cite{esterel.primer}:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''.}
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 4--9, in both 
implementations), given that Esterel's `$\|$' and \CEU's \code{par/and} 
constructs are equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (ln. 3--10), which, in this case, serves the same purpose of 
\CEU's \code{par/or} (ln. 3--12):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
// ESTEREL
loop
   abort
      [
         await A
      $\|$
         await B
      ];
      emit O
   when R
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
// CEU
loop do
   par/or do
      par/and do
         await A;
      with
         await B;
      end
      emit O;
   with
      await R;
   end
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ A control specification implemented in Esterel and \CEU:
\emph{``Emit \code{O} after \code{A} and \code{B}, resetting each \code{R}.''}
\label{lst.abro}
}
\end{figure}

@@

Esterel makes no semantic distinctions between internal and external signals.
In particular, programs can emit multiple external input signals, which may 
coexist during a reaction.
%
In \CEU, a reaction starts from the occurrence of an external input event and 
programs cannot emit inputs.
Therefore, the occurring input event is unique during the entire reaction, 
resulting in intrinsic queue-based handling.
%
In contrast, programs can emit internal events but these follow a stack-based 
execution policy, similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel and \CEU.
%
In Esterel, when \code{A} occurs, \code{B} is emitted (ln.  5--6) and both 
events become active, resulting in the invocation of \code{f()} and \code{g()} 
in no particular order.
%
In \CEU, the occurrence of \code{A} makes the program behave as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (ln. 5), emits \code{b}, and pauses.
\item 2nd trail awakes (ln. 9), calls \code{\_g()}, and terminates.
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates.
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

\begin{figure}[b]
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
// ESTEREL
input A;    // external
signal B;   // internal
[[
    await A;
    emit B;
    call f();
$\|$
    await B;
    call g();
]]
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
// CEU
input void A;  // external (in uppercase)
event void b;  // internal (in lowercase)
par/and do
    await A;
    emit b;
    _f();
with
    await b;
    _g();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Internal signals (events) in Esterel and \CEU: similar syntax, but 
different semantics.
\label{lst.prints}
}
\end{figure}

@@

Concurrency in \CEU is characterized when two or more trail segments in 
parallel execute during the same reaction.
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
%
In the program in the left of Figure~\ref{lst.shared}, the two assignments to 
\code{x} (ln. 5,8) can never run concurrently, because each trail segment 
reacts to different input events (ln. 4,7), which cannot occur simultaneously 
(according to Section~\ref{sec.ceu.ints}).
%
However, the program in the right is non-deterministic, because the two 
assignments to \code{y} (ln. 5,8) occur in the same reaction to input \code{A} 
(ln. 4,7).

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A, B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
the code in the left is safe because the trails access \code{x} atomically in 
different reactions;
the code in the right is unsafe because both trails access \code{y} in the same 
reaction.
\label{lst.shared}
}
\end{figure}

@@

Regardless of the temporal analysis of \CEU, when multiple trails are active 
during the same reaction, they are scheduled in the order they appear in the 
program source code.
%
Therefore, even though the program in the right of Figure~\ref{lst.shared} is 
suspicious, the assignments to \code{y} are both atomic and deterministic, 
i.e., after the reaction to \code{A} terminates, the value of \code{y} is $4$ 
(\code{(1+1)*2}).
%
On the one hand, enforcing an execution order for concurrent operations may 
seen arbitrary and also precludes true parallelism.
On the other hand, it provides a priority scheme for trails, and makes 
shared-memory concurrency more tractable.
%
For constrained embedded development, we believe that deterministic 
shared-memory concurrency is beneficial, given the extensive use of memory 
mapped ports for \emph{I/O} and the lack of hardware support for real 
parallelism.
Other synchronous embedded languages, such as \emph{SOL}~\cite{wsn.sol} and 
\emph{PRET-C}~\cite{rp.pretc}, made a similar design choice.
%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
%parallelism is inherent.

Figure~\ref{lst.det} compares the two syntactically equivalent code fragments 
in Esterel and \CEU to summarize the semantic difference regarding 
(non-)determinism.
%
Even though the program in \CEU executes deterministically, the compiler still 
issues a warning, because an apparently innocuous reordering of trails modifies 
the semantics of the program.
%
Note that in Esterel multiple external events can coexist in the same reaction, 
which disallows a similar temporal analysis.

\begin{figure}[b]
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[mathescape=true]
// ESTEREL
input A;
[
    await A;
    call f1();
$\|$
    await A;
    call f2();
];
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
// CEU
input void A;
par/and do
    await A;
    _f1();
with
    await A;
    _f2();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ In Esterel, the execution order between \code{f1} and \code{f2} is 
unspecified, whereas in \CEU, \code{\_f1} executes before \code{\_f2} due to 
deterministic scheduling based on lexical order.
\label{lst.det}
}
\end{figure}

@@

\begin{figure}
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void STOP, RETRANSMIT, SEND_ACK;
output _pkt_t* SEND_ENQUEUE, SEND_CANCEL;
par/or do
    await STOP;
with
    loop do
        par/or do
            await RETRANSMIT;
        with
            par/and do
                await 1min;
            with
                var _pkt_t buffer;
                <fill-buffer-info>
                emit SEND_ENQUEUE => &buffer;
                await SEND_ACK;
            end
        end
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em,firstnumber=12]
<...>
    var _pkt_t buffer;
    <fill-buffer-info>
    finalize
        emit SEND_ENQUEUE => &buffer;
    with
        emit SEND_CANCEL => &buffer;
    end
    await SEND_ACK;
<...>
\end{lstlisting}
\end{minipage}
%\rule{8.5cm}{0.37pt}
\caption{The unsafe network protocol in the left, which does not compile, is 
extended with a finalization clause in the right to handle abortion properly.
%\caption{ Parallel compositions can describe complex state machines.
\label{lst.fin}
}
\end{figure}

To illustrate the risks related to abortion, consider the unsafe example in the 
left of Figure~\ref{lst.fin}, which does not compile in \CEU.
It describes the state machine of the data collection protocol \emph{CTP} for 
sensor networks~\cite{wsn.ctp}.
%
The input and output events represent the external interface of the protocol 
(ln. 1--2).
%
The protocol has to transmit a packet every minute (with \code{SEND\_ENQUEUE}), 
unless it receives a \code{RETRANSMIT} request to immediately re-transmit it, 
or a \code{STOP} request to terminate.
%
The protocol is implemented with two main trails:
one simply monitors the stopping event (ln. 4);
the other periodically transmits the packet (ln. 6--19).
%
The periodic transmission is a loop that starts two other trails (ln. 7--18):
one handles the immediate retransmission request (ln. 8);
the other transmits the packet and waits for a confirmation (ln. 10--17).
%
%
The actual transmission (ln. 13--16) is enclosed with a \code{par/and} that 
takes at least one minute before looping (ln. 11), in accordance with the 
specification.
Note that the \code{emit SEND\_ENQUEUE} (ln. 15) is \emph{asynchronous}, 
handing to the radio driver a pointer to the lexically-scoped packet (ln. 13).
The driver makes the transmission in the background, holding the packet until 
it signals the application with a \code{SEND\_ACK} event to acknowledge 
completion (ln. 16).
%
At any time, the client may request a retransmission (ln. 8), which terminates 
the \code{par/or} (ln. 7) and restarts the loop (ln. 6), but does not abort the 
ongoing transmission initiated with the \code{emit SEND\_ENQUEUE} (ln. 15).
%
The client may also request to stop the whole protocol at any time (ln. 4).
%
Therefore, if the sending trail is aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope (ln. 13), 
leaving behind a \emph{dangling pointer} in the radio driver, which will 
possibly transmit corrupted data.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

The unsafe example in the left of Figure~\ref{lst.fin} does not compile because 
\CEU tracks the interaction of \code{par/or} compositions with local variables 
and stateful output events calls in order to preserve safe abortion of trails.
%
%Finalization clauses are fundamental to preserve the orthogonality of 
%\code{par/or} compositions in SSRP.
%
\CEU enforces a \emph{finalization} clause to accompany the output request.
The code in the right of Figure~\ref{lst.fin} properly cancels the packet 
transmission when the block of \code{buffer} goes out of scope, i.e., the 
finalization clause (after the \code{with}, ln. 18) executes automatically on 
external abortion.%
\footnote{
Note that the compiler only enforces the programmer to write the finalization 
clause, but cannot check if it actually handles the resource properly.
}
%
%Note also that \CEU environments rely on \C libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.

@@

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
// ESTEREL
abort
    await A;
    call f();
when A;

\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
// CEU
par/or do
    await A;
with
    await A;
    _f();
end
\end{lstlisting}
\end{minipage}
%
%\rule{8.4cm}{0.37pt}
\caption{ Strong abortion in Esterel and \CEU. %\newline
\label{lst.abortion}
}
\end{figure}
\end{comment}

@@

The \code{await} statement of \CEU supports wall-clock time and handles deltas 
automatically, resulting in more robust applications.
For the example in the left of Figure~\ref{lst.timers}, suppose that after the 
first \code{await} request, the underlying system gets busy and takes 15ms to 
check for expiring awaits.
The \CEU scheduler will notice that the \code{await 10ms} has not only already 
expired, but is delayed with \code{delta=5ms}.
Then, the awaiting trail awakes, sets \code{v=1}, and invokes \code{await 1ms}.
As the current delta is higher than the requested timeout (i.e. $5ms > 1ms$), 
the trail is rescheduled for execution, now with \code{delta=4ms}.

\begin{figure}
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}
var int v;
await 10ms;
v = 1;
await 1ms;
v = 2;
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.51\linewidth}
\begin{lstlisting}
par/or do
    await 10ms;
    <...>         // any non-awaiting sequence
    await  1ms;
    v = 1;
with
    await 12ms;
    v = 2;
end
\end{lstlisting}
\end{minipage}
%\rule{8.5cm}{0.37pt}
\caption{ First-class timers in \CEU.
\label{lst.timers}
}
\end{figure}

\CEU also takes into account the fact that time is a physical quantity that can 
be added and compared.
For instance, for the program in the right of Figure~\ref{lst.timers}, although 
the scheduler cannot guarantee that the first trail terminates exactly in 11ms, 
it can at least ensure that the program always terminates with \code{v=1}.
%
Given that any non-awaiting sequence is considered to take no time in the 
synchronous model, the first trail is guaranteed to terminate before the second 
trail, because $10+1 < 12$.
%
A similar program in a language without first-class support for timers would 
depend on the execution timings for the code marked as \code{<...>}, making the 
reasoning about the execution behavior more difficult.

@@

The concrete \code{await} and \code{emit} primitives support communication of 
values between them.
In the two-step translation of Figure~\ref{lst.map.emit.await}, we start with 
the concrete program in \CEU, which communicates the value $1$ between the 
\code{emit} and \code{await} in parallel (left-most code).
In the intermediate translation, we include the shared variable \code{e\_} to 
hold the value being communicated between the two trails in order to simplify 
the \code{emit}.
Finally, we convert the program into the equivalent in the abstract syntax, 
translating side-effect statements into $mem$ expressions.
External events require a similar translation, i.e., each external event has a 
corresponding variable that is explicitly set by the environment before each 
reaction.

\begin{figure}[t]
\begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}
par/or do
  <...>
  emit e => 1;
with
  v = await e;
  _printf("%d\n",v);
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}
par/or do
  <...>
  e_ = 1;
  emit e;
with
  await e;
  v = e_;
  _printf("%d\n",v);
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}
<...> ; mem ; emit(e)
        or
await(e) ; mem ; mem
\end{lstlisting}
\end{minipage}
\caption{
Two-step translation from concrete to abstract \code{emit} and \code{await} 
expressions.
The concrete code in the left communicates the value \code{1} from the 
\code{emit} to the \code{await}.
The abstract code uses a shared variable to hold the value.
\label{lst.map.emit.await}
}
\end{figure}

@@

The event \code{DT} has an associated variable \code{DT\_} carrying the 
wall-clock time elapsed between two occurrences in sequence, as depicted by the 
two-step translation of Figure~\ref{lst.map.timers.2}.
In the concrete program in the left, the variable \code{dt} holds the residual 
delta time (as described in Section~\ref{sec.ceu.timers}) after awaking from 
the timer.
In the first step of the translation, we expand the \code{await 10ms} to a 
\code{loop} that decrements the elapsed number of microseconds for each 
occurrence of \code{DT}.
When the variable \code{tot} reaches zero, we escape the \code{loop} setting 
the variable \code{dt} to contain the appropriate delta.
In the last step, we convert the program to the abstract syntax.

\begin{figure}
\begin{minipage}[t]{0.30\linewidth}
\begin{lstlisting}
dt = await 10ms;
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.37\linewidth}
\begin{lstlisting}
var int tot = 10000; // 10ms
loop do
    await DT;
    tot = tot - DT_;
    if tot <= 0 then
        dt = -tot;
        break;
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.30\linewidth}
\begin{lstlisting}
mem;
loop(
    await(DT);
    mem;
    if mem then
        mem;
        break
    else
        nop
)
\end{lstlisting}
\end{minipage}
\caption{
Two-step translation from concrete to abstract timer.
\label{lst.map.timers.2}
}
\end{figure}

@@

// CODE-1
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.26\linewidth}
\begin{lstlisting}
par/or do
  var int* ptr = <...>;
  await A;
  _hold(ptr);
  await B;
with
  { fin
      _release(ptr); }
end

>>>>

Regarding the similarities, Figure~\ref{lst.abro} shows side-by-side the 
implementations in Esterel (a) and \CEU (b) for the following control 
specification~\cite{esterel.primer}:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''.}
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 4--9, in both 
implementations), given that Esterel's `$\|$' and \CEU's \code{par/and} 
constructs are equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (ln. 3--10), which, in this case, serves the same purpose of 
\CEU's \code{par/or} (ln. 3--12):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
loop
   abort
      [
         await A
      $\|$
         await B
      ];
      emit O
   when R
end

.
\end{lstlisting}
\centering\small{(a) Esterel}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
loop do
   par/or do
      par/and do
         await A;
      with
         await B;
      end
      emit O;
   with
      await R;
   end
end
\end{lstlisting}
\centering\small{(b) \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ A control specification implemented in Esterel and \CEU:
\emph{``Emit \code{O} after \code{A} and \code{B}, resetting each \code{R}.''}
\label{lst.abro}
}
\end{figure}

@@

Esterel makes no semantic distinctions between internal and external signals.
In particular, programs can emit multiple external input signals, which may 
coexist during a reaction.
%
In \CEU, a reaction starts from the occurrence of an external input event and 
programs cannot emit inputs.
Therefore, the occurring input event is unique during the entire reaction, 
resulting in intrinsic queue-based handling.
%
In contrast, programs can emit internal events but these follow a stack-based 
execution policy, similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel (a) and \CEU (b).
%
In Esterel, when \code{A} occurs, \code{B} is emitted (ln.  5--6) and both 
events become active, resulting in the invocation of \code{f()} and \code{g()} 
in no particular order.
%
In \CEU, the occurrence of \code{A} makes the program behave as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (ln. 5), emits \code{b}, and pauses.
\item 2nd trail awakes (ln. 9), calls \code{\_g()}, and terminates.
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates.
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

\begin{figure}[b]
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
input A;    // external
signal B;   // internal
[[
    await A;
    emit B;
    call f();
$\|$
    await B;
    call g();
]]
\end{lstlisting}
\centering\small{(a) Esterel}
\end{minipage}
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
input void A;  // external (in uppercase)
event void b;  // internal (in lowercase)
par/and do
    await A;
    emit b;
    _f();
with
    await b;
    _g();
end
\end{lstlisting}
\centering\small{(b) \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Internal signals (events) in Esterel and \CEU: similar syntax, but 
different semantics.
\label{lst.prints}
}
\end{figure}

@@

Concurrency in \CEU is characterized when two or more trail segments in 
parallel execute during the same reaction.
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
%
In the example of Figure~\ref{lst.shared}.a, the two assignments to \code{x} 
(ln.  5,8) can never run concurrently, because each trail segment reacts to 
different input events (ln. 4,7), which cannot occur simultaneously (according 
to Section~\ref{sec.ceu.ints}).
%
However, the example of Figure~\ref{lst.shared}.b is non-deterministic, because 
the two assignments to \code{y} (ln. 5,8) occur in the same reaction to input 
\code{A} (ln. 4,7).

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A, B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\centering\small{(a)}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\centering\small{(b)}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
Example (a) is safe because the trails access \code{x} atomically in different 
reactions;
Example (b) is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure}

@@

Regardless of the temporal analysis of \CEU, when multiple trails are active 
during the same reaction, they are scheduled in the order they appear in the 
program source code.
%
Therefore, even though the example of Figure~\ref{lst.shared}.b is suspicious, 
the assignments to \code{y} are both atomic and deterministic, i.e., after the 
reaction to \code{A} terminates, the value of \code{y} is $4$ (\code{(1+1)*2}).
%
On the one hand, enforcing an execution order for concurrent operations may 
seen arbitrary and also precludes true parallelism.
On the other hand, it provides a priority scheme for trails, and makes 
shared-memory concurrency more tractable.
%
For constrained embedded development, we believe that deterministic 
shared-memory concurrency is beneficial, given the extensive use of memory 
mapped ports for \emph{I/O} and the lack of hardware support for real 
parallelism.
Other synchronous embedded languages, such as \emph{SOL}~\cite{wsn.sol} and 
\emph{PRET-C}~\cite{rp.pretc}, made a similar design choice.
%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
%parallelism is inherent.

Figure~\ref{lst.det} compares the two syntactically equivalent code fragments 
in Esterel (a) and \CEU (b) to summarize the semantic difference regarding 
(non-)determinism.
%
Even though the program in \CEU executes deterministically, the compiler still 
issues a warning, because an apparently innocuous reordering of trails modifies 
the semantics of the program.
%
Note that in Esterel multiple external events can coexist in the same reaction, 
which disallows a similar temporal analysis.

\begin{figure}[b]
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[mathescape=true]
input A;
[
    await A;
    call f1();
$\|$
    await A;
    call f2();
];
\end{lstlisting}
\centering\small{(a) Esterel}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
input void A;
par/and do
    await A;
    _f1();
with
    await A;
    _f2();
end
\end{lstlisting}
\centering\small{(b) \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ In Esterel, the execution order between \code{f1} and \code{f2} is 
unspecified, whereas in \CEU, \code{\_f1} executes before \code{\_f2} due to 
deterministic scheduling based on lexical order.
\label{lst.det}
}
\end{figure}

@@

\begin{figure}
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void STOP, RETRANSMIT, SEND_ACK;
output _pkt_t* SEND_ENQUEUE, SEND_CANCEL;
par/or do
    await STOP;
with
    loop do
        par/or do
            await RETRANSMIT;
        with
            par/and do
                await 1min;
            with
                var _pkt_t buffer;
                <fill-buffer-info>
                emit SEND_ENQUEUE => &buffer;
                await SEND_ACK;
            end
        end
    end
end
\end{lstlisting}
\centering\small{(a)}
\end{minipage}
%
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em,firstnumber=12]
<...>
    var _pkt_t buffer;
    <fill-buffer-info>
    finalize
        emit SEND_ENQUEUE => &buffer;
    with
        emit SEND_CANCEL => &buffer;
    end
    await SEND_ACK;
<...>

.
\end{lstlisting}
\centering\small{(b)}
\end{minipage}
%\rule{8.5cm}{0.37pt}
\caption{
(a) The unsafe network protocol does not compile.
(b) The finalization clause extends the protocol to handle abortion properly.
%\caption{ Parallel compositions can describe complex state machines.
\label{lst.fin}
}
\end{figure}

To illustrate the risks related to abortion, consider the unsafe example of 
Figure~\ref{lst.fin}.a, which does not compile in \CEU.
It describes the state machine of the data collection protocol \emph{CTP} for 
sensor networks~\cite{wsn.ctp}.
%
The input and output events represent the external interface of the protocol 
(ln. 1--2).
%
The protocol has to transmit a packet every minute (with \code{SEND\_ENQUEUE}), 
unless it receives a \code{RETRANSMIT} request to immediately re-transmit it, 
or a \code{STOP} request to terminate.
%
The protocol is implemented with two main trails:
one simply monitors the stopping event (ln. 4);
the other periodically transmits the packet (ln. 6--19).
%
The periodic transmission is a loop that starts two other trails (ln. 7--18):
one handles the immediate retransmission request (ln. 8);
the other transmits the packet and waits for a confirmation (ln. 10--17).
%
%
The actual transmission (ln. 13--16) is enclosed with a \code{par/and} that 
takes at least one minute before looping (ln. 11), in accordance with the 
specification.
Note that the \code{emit SEND\_ENQUEUE} (ln. 15) is \emph{asynchronous}, 
handing to the radio driver a pointer to the lexically-scoped packet (ln. 13).
The driver makes the transmission in the background, holding the packet until 
it signals the application with a \code{SEND\_ACK} event to acknowledge 
completion (ln. 16).
%
At any time, the client may request a retransmission (ln. 8), which terminates 
the \code{par/or} (ln. 7) and restarts the loop (ln. 6), but does not abort the 
ongoing transmission initiated with the \code{emit SEND\_ENQUEUE} (ln. 15).
%
The client may also request to stop the whole protocol at any time (ln. 4).
%
Therefore, if the sending trail is aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope (ln. 13), 
leaving behind a \emph{dangling pointer} in the radio driver, which will 
possibly transmit corrupted data.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

The unsafe example of Figure~\ref{lst.fin}.a does not compile because \CEU 
tracks the interaction of \code{par/or} compositions with local variables and 
stateful output events calls in order to preserve safe abortion of trails.
%
%Finalization clauses are fundamental to preserve the orthogonality of 
%\code{par/or} compositions in SSRP.
%
\CEU enforces a \emph{finalization} clause to accompany the output request.
The code in Figure~\ref{lst.fin}.b properly cancels the packet transmission 
when the block of \code{buffer} goes out of scope, i.e., the finalization 
clause (after the \code{with}, ln. 18) executes automatically on external 
abortion.%
\footnote{
Note that the compiler only enforces the programmer to write the finalization 
clause, but cannot check if it actually handles the resource properly.
}
%
%Note also that \CEU environments rely on \C libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.

@@

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
abort
    await A;
    call f();
when A;

\end{lstlisting}
\centering\small{(a) Esterel}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
par/or do
    await A;
with
    await A;
    _f();
end
\end{lstlisting}
\centering\small{(b) \CEU}
\end{minipage}
%
%\rule{8.4cm}{0.37pt}
\caption{ Strong abortion in Esterel and \CEU. %\newline
\label{lst.abortion}
}
\end{figure}
\end{comment}

@@

The \code{await} statement of \CEU supports wall-clock time and handles deltas 
automatically, resulting in more robust applications.
For the example of Figure~\ref{lst.timers}.a, suppose that after the first 
\code{await} request, the underlying system gets busy and takes 15ms to check 
for expiring awaits.
The \CEU scheduler will notice that the \code{await 10ms} has not only already 
expired, but is delayed with \code{delta=5ms}.
Then, the awaiting trail awakes, sets \code{v=1}, and invokes \code{await 1ms}.
As the current delta is higher than the requested timeout (i.e. $5ms > 1ms$), 
the trail is rescheduled for execution, now with \code{delta=4ms}.

\begin{figure}
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}
var int v;
await 10ms;
v = 1;
await 1ms;
v = 2;

.
\end{lstlisting}
\centering\small{(a)}
\end{minipage}
%
\begin{minipage}[t]{0.51\linewidth}
\begin{lstlisting}
par/or do
    await 10ms;
    <...>         // any non-awaiting sequence
    await  1ms;
    v = 1;
with
    await 12ms;
    v = 2;
end
\end{lstlisting}
\centering\small{(b)}
\end{minipage}
%\rule{8.5cm}{0.37pt}
\caption{ First-class timers in \CEU.
\label{lst.timers}
}
\end{figure}

\CEU also takes into account the fact that time is a physical quantity that can 
be added and compared.
For instance, for the example of Figure~\ref{lst.timers}.b, although the 
scheduler cannot guarantee that the first trail terminates exactly in 11ms, it 
can at least ensure that the program always terminates with \code{v=1}.
%
Given that any non-awaiting sequence is considered to take no time in the 
synchronous model, the first trail is guaranteed to terminate before the second 
trail, because $10+1 < 12$.
%
A similar program in a language without first-class support for timers would 
depend on the execution timings for the code marked as \code{<...>}, making the 
reasoning about the execution behavior more difficult.

@@

The concrete \code{await} and \code{emit} primitives support communication of 
values between them.
In the two-step translation of Figure~\ref{lst.map.emit.await}, we start with 
the concrete program in \CEU, which communicates the value $1$ between the 
\code{emit} and \code{await} in parallel (a).
In the intermediate translation (b), we include the shared variable \code{e\_} 
to hold the value being communicated between the two trails in order to 
simplify the \code{emit}.
Finally, we convert the program into the equivalent in the abstract syntax (c), 
translating side-effect statements into $mem$ expressions.
External events require a similar translation, i.e., each external event has a 
corresponding variable that is explicitly set by the environment before each 
reaction.

\begin{figure}[t]
\begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}
par/or do
  <...>
  emit e => 1;
with
  v = await e;
  _printf("%d\n",v);
end

.
\end{lstlisting}
\centering\small{(a)}
\end{minipage}
%
\begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}
par/or do
  <...>
  e_ = 1;
  emit e;
with
  await e;
  v = e_;
  _printf("%d\n",v);
end
\end{lstlisting}
\centering\small{(b)}
\end{minipage}
%
\begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}
<...> ; mem ; emit(e)
        or
await(e) ; mem ; mem

.
\end{lstlisting}
\centering\small{(c)}
\end{minipage}
\caption{
Two-step translation from concrete to abstract \code{emit} and \code{await} 
expressions.
The concrete code in (a) communicates the value \code{1} from the \code{emit} 
to the \code{await}.
The abstract code in (c) uses a shared variable to hold the value.
\label{lst.map.emit.await}
}
\end{figure}

@@

The event \code{DT} has an associated variable \code{DT\_} carrying the 
wall-clock time elapsed between two occurrences in sequence, as depicted by the 
two-step translation of Figure~\ref{lst.map.timers.2}.
In the concrete program (a), the variable \code{dt} holds the residual delta 
time (as described in Section~\ref{sec.ceu.timers}) after awaking from the 
timer.
In the first step of the translation (b), we expand the \code{await 10ms} to a 
\code{loop} that decrements the elapsed number of microseconds for each 
occurrence of \code{DT}.
When the variable \code{tot} reaches zero, we escape the \code{loop} setting 
the variable \code{dt} to contain the appropriate delta.
In the last step (c), we convert the program to the abstract syntax.

\begin{figure}
\begin{minipage}[t]{0.30\linewidth}
\begin{lstlisting}
dt = await 10ms;

.
\end{lstlisting}
\centering\small{(a)}
\end{minipage}
%
\begin{minipage}[t]{0.37\linewidth}
\begin{lstlisting}
var int tot = 10000; // 10ms
loop do
    await DT;
    tot = tot - DT_;
    if tot <= 0 then
        dt = -tot;
        break;
    end
end
.
\end{lstlisting}
\centering\small{(b)}
\end{minipage}
%
\begin{minipage}[t]{0.30\linewidth}
\begin{lstlisting}
mem;
loop(
    await(DT);
    mem;
    if mem then
        mem;
        break
    else
        nop
)
\end{lstlisting}
\centering\small{(c)}
\end{minipage}
\caption{
Two-step translation from concrete to abstract timer.
\label{lst.map.timers.2}
}
\end{figure}

@@

.
\end{lstlisting}
\centering\small{(a)}
\end{minipage}
%
\begin{minipage}[t]{0.26\linewidth}
\begin{lstlisting}
par/or do
  var int* ptr = <...>;
  await A;
  _hold(ptr);
  await B;
with
  { fin
      _release(ptr); }
end
